---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# validatoR

<!-- badges: start -->
[![GitHub Release](https://img.shields.io/github/v/release/bonifazi/validatoR?include_prereleases)](https://github.com/bonifazi/validatoR/releases) [![Licence: MIT](https://img.shields.io/badge/Licence-MIT-green)](https://github.com/bonifazi/validatoR/blob/main/LICENSE.md) [![R-CMD-check](https://github.com/bonifazi/validatoR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bonifazi/validatoR/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/bonifazi/validatoR/graph/badge.svg)](https://app.codecov.io/gh/bonifazi/validatoR)
<!-- badges: end -->

validatoR validates (genomic) prediction models in animal breeding. You give it
predictions, such as (genomic) EBVs, and the thing they're meant to predict,
such as pre-corrected phenotypes, true breeding values or EBVs from a later
evaluation. It returns the statistics you need to judge them: correlation,
slope, intercept, mean difference and, if you supply the heritability, accuracy.
Standard errors come from a bootstrap via the boot package.

## Installation

You can install the development version of validatoR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("bonifazi/validatoR")
```

## Example

`toy_validation` ships with the package. It holds 2000 animals with EBVs from a
partial and a whole evaluation, built so the true answers are known.

```{r example}
library(validatoR)

res <- validate_prediction(
  prediction = toy_validation$partial,
  target     = toy_validation$whole,
  h2         = 0.8
)
res$stats
```

Add `bootstrap = TRUE` to get standard errors, and `plot = TRUE` for a
scatter plot of target against prediction.

```{r example-boot}
set.seed(1)
res_boot <- validate_prediction(
  toy_validation$partial, toy_validation$whole,
  bootstrap = TRUE, n_boot = 200, plot = TRUE
)
res_boot$stats
res_boot$plot
```
