---
title: "Getting started with validatoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with validatoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center"
)
```

This vignette shows what validatoR can do, on a small simulated data set. It
assumes you know what an EBV is and how a genetic evaluation is run, and it
explains only what is specific to the package.

## Which function to use

| You have                                                                                                                                            | You want                                                                                                                                                        | Use                      |
| :-------------------------------------------------------------------------------------------------------------------------------------------------- | :-------------------------------------------------------------------------------------------------------------------------------------------------------------- | :----------------------- |
| EBVs expressed on different bases                                                                                                                   | EBVs on a common base, before you compare or validate them                                                                                                      | `rebase_ebv()`           |
| Two evaluations of the same animals: a partial one, which leaves out the records of the validation animals, and a whole one, which uses all records | The level bias, the dispersion bias, the accuracy of the partial EBVs (the LR method), and the ratio of accuracies                                              | `validate_lr()`          |
| Two evaluations as above, and groups of animals such as cohorts, herds or sexes                                                                     | The same statistics for each group, in one table                                                                                                                | `validate_lr_by_group()` |
| Predictions and a target to validate them with: pre-corrected phenotypes, true breeding values from a simulation, or other EBVs                     | The correlation, the regression of the target on the prediction, the mean difference and the MSE, and the accuracy when the target is a pre-corrected phenotype | `validate_prediction()`  |

The validation animals are the animals whose EBVs you validate, for example the
youngest generation. The LR method assumes that the evaluation accounts for
selection and that the validation set is large and diverse (see Further
reading). All three functions can also return a plot.

## The example data

```{r data}
library(validatoR)

head(toy_validation)
```

`toy_validation` has 2,000 simulated animals: their EBVs from a partial and a
whole evaluation, a pre-corrected phenotype simulated with a heritability of
0.3, a cohort (`group`) and the inbreeding coefficient. The simulation fixes the
LR statistics at exact values (level bias 0.75, dispersion bias 0.90 and rho
0.85, the correlation between the partial and the whole EBVs). The additive
genetic variance is 300.

## Rebasing EBVs

Rebase first when the EBVs you want to validate are on different bases, and
validate the rebased EBVs afterwards. `rebase_ebv()` subtracts the mean EBV of a
base population from the EBVs of all animals, so that the base population has a
mean of 0 in each EBV column. The base population must lie outside the
validation animals: the level bias of a validation is one of the quantities you
want to estimate, and rebasing on the validation animals themselves would remove
part of it. `rebase_ebv()` takes two data frames. The first has the ID in
column 1 and one column of EBVs per trait or evaluation, and the second has the
IDs of the base population.

```{r rebase}
# the EBVs to rebase
ebv <- toy_validation[, c("id", "partial", "whole")]
# the IDs of the base population
base_pop <- toy_validation[toy_validation$group == "cohort_1", "id", drop = FALSE]

rebased <- rebase_ebv(ebv, base_pop = base_pop)
head(rebased$rebased_ebv)
```

Cohort 1 only illustrates the call. In a real validation, the base population is
a group outside the validation animals, for example older animals. The result is
a list, and the data frame is `rebased$rebased_ebv`. Every animal in a column
moves by the same amount, so differences between animals do not change.

If you already know the values to subtract, use `constant_value` instead of
`base_pop`, with one number per EBV column, in the order of the columns. Here
100 is subtracted from `partial` and 90 from `whole`. With `plot = TRUE`, the
result also has `plots`, a named list with one plot per EBV column, named after
the columns of your data. So here the plot of the column called `partial` is
`plots$partial`, and with your own data the names follow your own columns. Each
plot shows the rebased EBVs against the original ones: the points stay on a line
of slope 1, shifted by the value subtracted.

```{r rebase-constant, fig.alt = "Scatter plot of the rebased EBVs against the original EBVs. The points lie on a straight line of slope 1, shifted down by the value subtracted, so the correlation and the slope stay 1. Dotted lines mark the mean of each."}
rebased_constant <- rebase_ebv(ebv, constant_value = c(100, 90), plot = TRUE)
head(rebased_constant$rebased_ebv, 3)
# the plot of the column called "partial"
rebased_constant$plots$partial
```

The examples below use the EBVs of `toy_validation` as they are, because its
two evaluations are already on the same base.

## The LR method

`validate_lr()` takes one data frame per evaluation, with the animal ID in
column 1 and the EBV in column 2, and matches animals by ID.

```{r lr-input}
# get the partial EBV
partial <- toy_validation[, c("id", "partial")]
# get the whole EBV
whole <- toy_validation[, c("id", "whole")]

res <- validate_lr(partial, whole)
res$stats # view the statistics
```

The result is a list with `stats`, the table above, and `plot`. The rows of
`stats` are:

| Row               | Meaning                                                                                                                                         | Expectation                                                                                                                                                                            |
| :---------------- | :---------------------------------------------------------------------------------------------------------------------------------------------- | :------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| `n`               | Number of validation animals: the animals found in both the partial and the whole evaluation, or the animals of `val_group` when it is provided |                                                                                                                                                                                        |
| `level_bias`      | Mean of the partial EBVs minus mean of the whole EBVs                                                                                           | Expected to be 0 if the partial EBVs are unbiased                                                                                                                                      |
| `dispersion_bias` | Slope of the regression of the whole EBVs on the partial EBVs                                                                                   | Expected to be 1 if the partial EBVs are unbiased. Below 1 the partial EBVs are over-dispersed relative to the whole EBVs (or the whole EBVs are under-dispersed), above 1 the reverse |
| `rho`             | Correlation between the partial and the whole EBVs, the ratio of their accuracies                                                               | rho = 1 means no gain in accuracy when moving from the partial to the whole evaluation                                                                                                 |
| `inc_acc`         | `1 / rho`, the increase in accuracy obtained with the whole evaluation, as a ratio to the accuracy of the partial evaluation                    |                                                                                                                                                                                        |

In the example, the level bias is
`r round(res$stats["level_bias", "value"], 2)`, the dispersion
bias is `r round(res$stats["dispersion_bias", "value"], 2)`
and rho is `r round(res$stats["rho", "value"], 2)`, exactly the values that were
simulated. The partial EBVs are over-dispersed, as simulated.

To express `inc_acc` as the increase in accuracy relative to the partial
evaluation, in percentage, subtract 1 and multiply by 100 (Bonifazi et al.,
2022). Here that is `r round((res$stats["inc_acc", "value"] - 1) * 100, 1)`%.
For example, a rho of 0.80 gives an `inc_acc` of 1.25, a relative increase in
population accuracy of 25% when moving from the partial to the whole
evaluation.

Legarra and Reverter (2018) suggest checking the level bias and the dispersion
first, and comparing accuracies only for evaluations that are approximately
unbiased, because the accuracy statistics rely on unbiasedness. If possible, use
large validation sets of several hundred animals or more.

### Accuracy of the partial EBVs

With `var_a` and the inbreeding of the validation animals, `validate_lr()` adds
`accuracy_partial`, the population accuracy of the partial EBVs (the correlation
between true and estimated breeding values across the validation animals), and
`level_bias_in_GSD`, the level bias in genetic standard deviations. `var_a` is the
additive genetic variance. The inbreeding is a data frame with the ID in
column 1 and the coefficient in column 2, or `average_inbreeding`, the
average inbreeding, if you only have that.

```{r lr-accuracy}
# get the inbreeding coefficients
inbreeding <- toy_validation[, c("id", "inbreeding")]

res_acc <- validate_lr(partial, whole, var_a = 300, inbreeding = inbreeding)
res_acc$stats # view the statistics
```

### Standard errors

With `bootstrap = TRUE` the validation animals are resampled with replacement
and every statistic is recomputed in each resample. The `SE` column is the
standard deviation of each statistic across the resamples, and an approximate
95% interval is the estimate plus or minus two standard errors. The default is
10,000 resamples and here is 200 just to keep this vignette fast.

```{r lr-bootstrap}
# make the resamples reproducible
set.seed(1)
res_boot <- validate_lr(
  partial,
  whole,
  var_a = 300,
  inbreeding = inbreeding,
  bootstrap = TRUE,
  n_boot = 200
)
res_boot$stats
```

### Plots

`plot = TRUE` adds a scatter plot of the whole EBVs on the partial EBVs. Each
point is an animal. The grey line has slope 1, and the blue line is the
regression of the whole EBVs on the partial EBVs, so its slope is the
dispersion bias.

```{r lr-plot, fig.alt = "Scatter plot of the whole EBVs against the partial EBVs, with a grey line of slope 1 and a blue regression line."}
res_plot <- validate_lr(partial, whole, plot = TRUE)
res_plot$plot
```

With `var_a`, `plot_in_gsd = TRUE` draws both axes in genetic standard
deviations, and `plot_verbose = TRUE` adds the number of animals as a subtitle
and the main statistics as a caption.

```{r lr-plot-gsd, fig.alt = "Scatter plot of the whole against the partial EBVs with both axes in genetic standard deviations, with the number of animals as a subtitle and the main statistics as a caption."}
res_gsd <- validate_lr(
  partial,
  whole,
  var_a = 300,
  plot = TRUE,
  plot_in_gsd = TRUE,
  plot_verbose = TRUE
)
res_gsd$plot
```

### Choosing the validation animals

By default, `validate_lr()` uses all animals found in both data frames. Often
you want to validate only some animals, such as the youngest selection candidates.
To do that, pass `val_group`: a data frame with the IDs of the validation animals in column 1 and, optionally, a
label in column 2.

The statistics are computed once, on all the animals of `val_group` together.
Before you trust one set of statistics, it is reccomended to check that the validation group is homogeneous, that is, that its animals behave alike. To help with this, you can label the animals in column 2 of `val_group`, for example by cohort or by sex, then setting `plot_subgroups = TRUE` colours the points by that label, so you can see whether one subgroup differs from the others. If the group is not homogeneous, you may want to validate its subgroups separately. The next section shows how.

```{r lr-group, fig.alt = "Scatter plot of the whole against the partial EBVs for the animals of cohorts 1 and 2, with the points coloured by cohort."}
# the validation animals: the animals of cohorts 1 and 2, labelled by cohort
val_group <- toy_validation[
  toy_validation$group %in% c("cohort_1", "cohort_2"),
  c("id", "group")
]

res_group <- validate_lr(
  partial,
  whole,
  val_group = val_group,
  plot = TRUE,
  plot_subgroups = TRUE
)
res_group$stats
res_group$plot
```
In this example, the two cohorts behave alike.

### Validate several groups at once

Sometimes you want to monitor several groups of animals and see whether the
validation differs between them: all animals, the young animals, the genotyped
and the non-genotyped animals, different ages, sexes, or lines. A bias that is
hidden in the whole population may only show up in one of these groups.
`validate_lr_by_group()` runs the validation for each group in one call and
returns the results of all groups in one table, so you do not have to repeat
`validate_lr()` for each group yourself.

It takes the same arguments as `validate_lr()`, plus a table, `groups`. This table contains the animal IDs in column 1 and one column per each group you want to validate. There are two types of columns you can define: logical and text. A logical column has `TRUE` for the animals in the group. A text column has a label for the animals in the group and `NA` for the rest. Note that an animal can be in several groups.

In this example, we define two logical columns, one for each cohort, and a text column with the cohort labels (either cohort_1 or cohort_2).

```{r lr-by-group}
# one column per group: two cohorts as logical columns, and the cohorts as labels
groups <- data.frame(
  id = toy_validation$id,
  cohort_1 = toy_validation$group == "cohort_1",
  cohort_2 = toy_validation$group == "cohort_2",
  cohort = toy_validation$group
)

# run the validation for all groups at once
res_by_group <- validate_lr_by_group(partial, whole, groups, var_a = 300)
```

The messages show each group as it runs and a summary at the end. The result
has a `groups` table, with one row for each group, and the statistics of all
groups in one long table:

```{r lr-by-group-tables}
res_by_group$groups
res_by_group$stats[res_by_group$stats$statistic == "rho", ]
```

`n_provided` is the number of IDs listed in the group, and `n` is the number of
animals used, which are the listed animals found in both evaluations. A group
that cannot be validated, for example with fewer than 3 animals, does not stop
the others from being validated. Its status is "failed" and its message says why.

The first group, `all`, is every animal found in both evaluations, whether or
not it is in a group. Its numbers are those of a plain `validate_lr()` call.
Use `include_all = FALSE` to leave it out.

With `plot = TRUE`, `plots` has one plot for each group. With
`plot_subgroups = TRUE`, the points of a text column are coloured by its labels.
Because the plots are in a list, you can assemble them yourself, for example
with `patchwork::wrap_plots(res$plots)`.

```{r lr-by-group-plots, fig.alt = "Scatter plot of the whole against the partial EBVs for all animals, with the points coloured by cohort."}
res_plots <- validate_lr_by_group(
  partial,
  whole,
  groups,
  plot = TRUE,
  plot_subgroups = TRUE,
  verbose = FALSE
)
res_plots$plots$cohort
```

With `split_labels = TRUE`, each label of a text column also becomes a group of
its own, named `cohort: cohort_1`, `cohort: cohort_2`, and so on.

## Predictions against a target

`validate_prediction()` compares two numeric vectors of the same length and
order: the first prediction belongs to the first target, and so on.

```{r prediction}
res_pred <- validate_prediction(
  prediction = toy_validation$partial,
  target = toy_validation$pheno,
  h2 = 0.3
)
res_pred$stats
```

| Row           | Meaning                                                                             |
| :------------ | :---------------------------------------------------------------------------------- |
| `n`           | Number of pairs                                                                     |
| `correlation` | Correlation between prediction and target                                           |
| `slope`       | Slope of the regression of the target on the prediction. 1 means no dispersion bias |
| `intercept`   | Intercept of that regression                                                        |
| `mean_diff`   | Mean prediction minus mean target, the level bias                                   |
| `mse`, `rmse` | Mean squared error and its square root, from `prediction - target`                  |
| `accuracy`    | Correlation divided by `sqrt(h2)`, only when `h2` is provided                       |

The accuracy is valid when the target is a pre-corrected phenotype that was not
used in the evaluation that produced the predictions. Here it is
`r round(res_pred$stats["accuracy", "value"], 2)`, the value that was simulated.
`h2` should be the heritability of the validation population. After selection it
is lower than in the base population, and the base-population value then
underestimates the accuracy. Estimates from phenotypes, including the slope, are
also sensitive to errors in the heritability and in the pre-correction for fixed
effects (Legarra and Reverter, 2018).

`mse` and `rmse` are only meaningful when prediction and target are on the same
scale, for example a partial and a whole EBV, or EBVs and true breeding values.
With `var_a`, `rmse_in_GSD` gives the RMSE in genetic standard deviations. This
function also takes `bootstrap = TRUE` and `plot = TRUE`, which draws the target
against the prediction:

```{r prediction-rmse, fig.alt = "Scatter plot of the target on the prediction, with a grey line of slope 1 and a blue regression line."}
res_rmse <- validate_prediction(
  prediction = toy_validation$partial,
  target = toy_validation$whole,
  var_a = 300,
  plot = TRUE
)
res_rmse$stats
res_rmse$plot
```

As in `validate_lr()`, `plot_subgroups = TRUE` colours the points by group. Here
`group` is a vector with one label per animal, in the same order as the data.
The statistics do not change.

```{r prediction-group, fig.alt = "Scatter plot of the phenotype on the partial EBVs, with the points coloured by cohort."}
res_pred_group <- validate_prediction(
  prediction = toy_validation$partial,
  target = toy_validation$pheno,
  h2 = 0.3,
  group = toy_validation$group,
  plot = TRUE,
  plot_subgroups = TRUE
)
res_pred_group$plot
```

## Messages and errors

The functions stop with a message when an input cannot be used. `validate_lr()`
leaves out animals found in only one of the two evaluations and reports how
many:

```{r message-example}
res_left_out <- validate_lr(partial[1:1990, ], whole)
```

## Further reading

Each function has a help page that describes every argument: `?rebase_ebv`,
`?validate_lr`, `?validate_lr_by_group` and `?validate_prediction`.

The LR method is described in Legarra, A. and Reverter, A. (2018). Semi-parametric
estimates of population accuracy and bias of predictions of breeding values and
future phenotypes using the LR method. *Genetics Selection Evolution* 50:53.
<https://doi.org/10.1186/s12711-018-0426-6>

The increase in accuracy as a percentage is described in Bonifazi, R., Calus,
M. P. L., ten Napel, J., Veerkamp, R. F., Michenet, A., Savoia, S., Cromie, A.
and Vandenplas, J. (2022). International single-step SNPBLUP beef cattle
evaluations for Limousin weaning weight. *Genetics Selection Evolution* 54:57.
<https://doi.org/10.1186/s12711-022-00748-0>