---
title: "Getting started with validatoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with validatoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center"
)
```

This vignette shows what validatoR can do, on a small simulated data set. It
assumes you know what an EBV is and how a genetic evaluation is run, and it
explains only what is specific to the package.

## Which function to use

| You have | You want | Use |
|:---------|:---------|:----|
| EBVs expressed on different bases | EBVs on a common base, before you compare or validate them | `rebase_ebv()` |
| Two evaluations of the same animals: a partial one, which leaves out the records of the validation animals, and a whole one, which uses all records | The level bias, the dispersion bias, the accuracy of the partial EBVs (the LR method), and the ratio of accuracies | `validate_lr()` |
| Predictions and a target to validate them with: pre-corrected phenotypes, true breeding values from a simulation, or other EBVs | The correlation, the regression of the target on the prediction, the mean difference and the MSE, and the accuracy when the target is a pre-corrected phenotype | `validate_prediction()` |

The validation animals are the animals whose EBVs you validate, for example the
youngest generation. The LR method assumes that the evaluation accounts for
selection and that the validation set is large and diverse (see Further
reading). All three functions can also return a plot.

## The example data

```{r data}
library(validatoR)

head(toy_validation)
```

`toy_validation` has 2,000 simulated animals: their EBVs from a partial and a
whole evaluation, a pre-corrected phenotype simulated with a heritability of
0.3, a cohort (`group`) and the inbreeding coefficient. The simulation fixes the
LR statistics at exact values (level bias 0.75, dispersion bias 0.90 and rho
0.85, the correlation between the partial and the whole EBVs). The additive
genetic variance is 300.

## Rebasing EBVs

Rebase first when the EBVs you want to validate are on different bases, and
validate the rebased EBVs afterwards. `rebase_ebv()` subtracts the mean EBV of a
base population from the EBVs of all animals, so that the base population has a
mean of 0 in each EBV column. The base population must lie outside the
validation animals: the level bias of a validation is one of the quantities you
want to estimate, and rebasing on the validation animals themselves would remove
part of it. `rebase_ebv()` takes two data frames. The first has the ID in
column 1 and one column of EBVs per trait or evaluation, and the second has the
IDs of the base population.

```{r rebase}
# the EBVs to rebase
ebv <- toy_validation[, c("id", "partial", "whole")]
# the IDs of the base population
base_pop <- toy_validation[toy_validation$group == "cohort_1", "id", drop = FALSE]

rebased <- rebase_ebv(ebv, base_pop = base_pop)
head(rebased$rebased_ebv)
```

Cohort 1 only illustrates the call. In a real validation, the base population is
a group outside the validation animals, for example older animals. The result is
a list, and the data frame is `rebased$rebased_ebv`. Every animal in a column
moves by the same amount, so differences between animals do not change.

If you already know the values to subtract, use `constant_value` instead of
`base_pop`, with one number per EBV column, in the order of the columns. Here
100 is subtracted from `partial` and 90 from `whole`. With `plot = TRUE`, the
result also has `plots`, a named list with one plot per EBV column, named after
the columns of your data. So here the plot of the column called `partial` is
`plots$partial`, and with your own data the names follow your own columns. Each
plot shows the rebased EBVs against the original ones: the points stay on a line
of slope 1, shifted by the value subtracted.

```{r rebase-constant}
rebased_constant <- rebase_ebv(ebv, constant_value = c(100, 90), plot = TRUE)
head(rebased_constant$rebased_ebv, 3)
# the plot of the column called "partial"
rebased_constant$plots$partial
```

The examples below use the EBVs of `toy_validation` as they are, because its
two evaluations are already on the same base.

## The LR method

`validate_lr()` takes one data frame per evaluation, with the animal ID in
column 1 and the EBV in column 2, and matches animals by ID.

```{r lr-input}
# get the partial EBV
partial <- toy_validation[, c("id", "partial")]
# get the whole EBV
whole <- toy_validation[, c("id", "whole")]

res <- validate_lr(partial, whole)
res$stats # view the statistics
```

The result is a list with `stats`, the table above, and `plot`. The rows of
`stats` are:

| Row | Meaning | Value if unbiased |
|:----|:--------|:------------------|
| `n` | Number of validation animals | |
| `level_bias` | Mean of the partial EBVs minus mean of the whole EBVs | 0 |
| `dispersion_bias` | Slope of the regression of the whole EBVs on the partial EBVs | 1. Below 1 the partial EBVs are over-dispersed relative to the whole EBVs (or the whole EBVs are under-dispersed), above 1 the reverse |
| `rho` | Correlation between the partial and the whole EBVs, the ratio of their accuracies | Close to 1 |
| `inc_acc` | `1 / rho`, the increase in accuracy obtained with the whole evaluation, as a ratio to the accuracy of the partial evaluation | 1 means no gain |

In the example, the level bias is
`r round(res$stats["level_bias", "value"], 2)`, the dispersion
bias is `r round(res$stats["dispersion_bias", "value"], 2)`
and rho is `r round(res$stats["rho", "value"], 2)`, exactly the values that were
simulated. The partial EBVs are over-dispersed, as simulated.

To express `inc_acc` as the increase in accuracy relative to the partial
evaluation, in percentage, subtract 1 and multiply by 100 (Bonifazi et al.,
2022). Here that is `r round((res$stats["inc_acc", "value"] - 1) * 100, 1)`%.
For example, a rho of 0.80 gives an `inc_acc` of 1.25, a relative increase in
population accuracy of 25% when moving from the partial to the whole
evaluation.

Legarra and Reverter (2018) suggest checking the level bias and the dispersion
first, and comparing accuracies only for evaluations that are approximately
unbiased, because the accuracy statistics rely on unbiasedness. In a small or
related validation set, the dispersion bias is expected to fall below 1 even when
the EBVs are correct (0.94 in the paper's example of 100 animals in half-sib
families of 10), so use large validation sets, several hundred animals or more.

### Accuracy of the partial EBVs

With `var_a` and the inbreeding of the validation animals, `validate_lr()` adds
`accuracy_partial`, the population accuracy of the partial EBVs (the correlation
between true and estimated breeding values across the validation animals), and
`level_bias_in_GSD`, the level bias in genetic standard deviations. `var_a` is the
additive genetic variance. The inbreeding is a data frame with the ID in
column 1 and the coefficient in column 2, or `average_inbreeding`, the
average inbreeding, if you only have that.

```{r lr-accuracy}
# get the inbreeding coefficients
inbreeding <- toy_validation[, c("id", "inbreeding")]

res_acc <- validate_lr(partial, whole, var_a = 300, inbreeding = inbreeding)
res_acc$stats # view the statistics
```

### Standard errors

With `bootstrap = TRUE` the validation animals are resampled with replacement
and every statistic is recomputed in each resample. The `SE` column is the
standard deviation of each statistic across the resamples, and an approximate
95% interval is the estimate plus or minus two standard errors. The default is
10000 resamples; 200 keep this vignette fast.

```{r lr-bootstrap}
# make the resamples reproducible
set.seed(1)
res_boot <- validate_lr(
  partial,
  whole,
  var_a = 300,
  inbreeding = inbreeding,
  bootstrap = TRUE,
  n_boot = 200
)
res_boot$stats
```

### Plots

`plot = TRUE` adds a scatter plot of the whole EBVs on the partial EBVs. Each
point is an animal. The grey line has slope 1, and the blue line is the
regression of the whole EBVs on the partial EBVs, so its slope is the
dispersion bias.

```{r lr-plot}
res_plot <- validate_lr(partial, whole, plot = TRUE)
res_plot$plot
```

With `var_a`, `plot_in_gsd = TRUE` draws both axes in genetic standard
deviations, and `plot_verbose = TRUE` adds the number of animals as a subtitle
and the main statistics as a caption.

```{r lr-plot-gsd}
res_gsd <- validate_lr(
  partial,
  whole,
  var_a = 300,
  plot = TRUE,
  plot_in_gsd = TRUE,
  plot_verbose = TRUE
)
res_gsd$plot
```

### Validation group

By default, `validate_lr()` uses all animals found in both data frames. To
validate a subset, such as the youngest animals, pass `val_group`: a data frame
with the IDs of the validation animals in column 1 and, optionally, a group
label in column 2. `plot_subgroups = TRUE` colours the points by that label,
which helps to spot a group that behaves differently.

```{r lr-group}
# the validation animals: the animals of cohorts 1 and 2, labelled by cohort
val_group <- toy_validation[
  toy_validation$group %in% c("cohort_1", "cohort_2"),
  c("id", "group")
]

res_group <- validate_lr(
  partial,
  whole,
  val_group = val_group,
  plot = TRUE,
  plot_subgroups = TRUE
)
res_group$stats
res_group$plot
```

## Predictions against a target

`validate_prediction()` compares two numeric vectors of the same length and
order: the first prediction belongs to the first target, and so on.

```{r prediction}
res_pred <- validate_prediction(
  prediction = toy_validation$partial,
  target = toy_validation$pheno,
  h2 = 0.3
)
res_pred$stats
```

| Row | Meaning |
|:----|:--------|
| `n` | Number of pairs |
| `correlation` | Correlation between prediction and target |
| `slope` | Slope of the regression of the target on the prediction. 1 means no dispersion bias |
| `intercept` | Intercept of that regression |
| `mean_diff` | Mean prediction minus mean target, the level bias |
| `mse`, `rmse` | Mean squared error and its square root, from `prediction - target` |
| `accuracy` | Correlation divided by `sqrt(h2)`, only when `h2` is provided |

The accuracy is valid when the target is a pre-corrected phenotype that was not
used in the evaluation that produced the predictions. Here it is
`r round(res_pred$stats["accuracy", "value"], 2)`, the value that was simulated.
`h2` should be the heritability of the validation population. After selection it
is lower than in the base population, and the base-population value then
underestimates the accuracy. Estimates from phenotypes, including the slope, are
also sensitive to errors in the heritability and in the pre-correction for fixed
effects (Legarra and Reverter, 2018).

`mse` and `rmse` are only meaningful when prediction and target are on the same
scale, for example a partial and a whole EBV, or EBVs and true breeding values.
With `var_a`, `rmse_in_GSD` gives the RMSE in genetic standard deviations. This
function also takes `bootstrap = TRUE` and `plot = TRUE`, which draws the target
against the prediction:

```{r prediction-rmse}
res_rmse <- validate_prediction(
  prediction = toy_validation$partial,
  target = toy_validation$whole,
  var_a = 300,
  plot = TRUE
)
res_rmse$stats
res_rmse$plot
```

As in `validate_lr()`, `plot_subgroups = TRUE` colours the points by group. Here
`group` is a vector with one label per animal, in the same order as the data.
The statistics do not change.

```{r prediction-group}
res_pred_group <- validate_prediction(
  prediction = toy_validation$partial,
  target = toy_validation$pheno,
  h2 = 0.3,
  group = toy_validation$group,
  plot = TRUE,
  plot_subgroups = TRUE
)
res_pred_group$plot
```

## Messages and errors

The functions stop with a message when an input cannot be used. `validate_lr()`
leaves out animals found in only one of the two evaluations and reports how
many:

```{r message-example}
res_left_out <- validate_lr(partial[1:1990, ], whole)
```

## Further reading

Each function has a help page that describes every argument: `?rebase_ebv`,
`?validate_lr` and `?validate_prediction`.

The LR method is described in Legarra, A. and Reverter, A. (2018). Semi-parametric
estimates of population accuracy and bias of predictions of breeding values and
future phenotypes using the LR method. *Genetics Selection Evolution* 50:53.
<https://doi.org/10.1186/s12711-018-0426-6>

The increase in accuracy as a percentage is described in Bonifazi, R., Calus,
M. P. L., ten Napel, J., Veerkamp, R. F., Michenet, A., Savoia, S., Cromie, A.
and Vandenplas, J. (2022). International single-step SNPBLUP beef cattle
evaluations for Limousin weaning weight. *Genetics Selection Evolution* 54:57.
<https://doi.org/10.1186/s12711-022-00748-0>